---
title: 'Corona Virus: Answering Questions I have'
author: "Ritwik Sinha"
date: "3/24/2020"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load libraries
library(data.table)
library(ggplot2)
library(dplyr)
library(rlang)
library(tidyverse)
library(tidyr)
library(knitr)

```

## Let's import the data

This is based on the data that is stored and updated daily on this Git Repository <https://github.com/CSSEGISandData/COVID-19>. Please clone this repo, and get the data locally, if you want to rerun this code. 

```{r message=FALSE, warning=FALSE, echo=FALSE}
dat_cases = read_csv("../../COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_confirmed_global.csv")

names(dat_cases)[1] = "State"
names(dat_cases)[2] = "Country"

for(i in 5:length(names(dat_cases))){
  names(dat_cases)[i] = paste("date", gsub("/", "_", names(dat_cases)[i]), sep="")
}

first_date = names(dat_cases)[5]
last_date = names(dat_cases)[length(names(dat_cases))]

top_countries = dat_cases %>% 
  group_by(Country) %>%
  summarize(total = sum(!!sym(last_date))) %>%
  arrange(desc(total)) %>%
  data.frame()
  
dat_cases_long = dat_cases %>%
  gather(Date, Value, !!sym(first_date):!!sym(last_date)) %>%
  data.frame()

dat_cases_long$State = NULL
dat_cases_long$Lat = NULL
dat_cases_long$Long = NULL

dat_cases_long = dat_cases_long %>%
  group_by(Country, Date) %>%
  summarise(total = sum(Value)) %>%
  data.frame()

dat_cases_long$Date  = as.Date(gsub("_", "-", gsub("date", "", dat_cases_long$Date)), "%m-%d-%y")

dat_cases_long = dat_cases_long %>%
  arrange(Country, Date)

dat_deaths = read_csv("../../COVID-19/csse_covid_19_data/csse_covid_19_time_series/time_series_covid19_deaths_global.csv")

names(dat_deaths)[1] = "State"
names(dat_deaths)[2] = "Country"

for(i in 5:length(names(dat_deaths))){
  names(dat_deaths)[i] = paste("date", gsub("/", "_", names(dat_deaths)[i]), sep="")
}

first_date = names(dat_deaths)[5]
last_date = names(dat_deaths)[length(names(dat_deaths))]

top_countries_deaths = dat_deaths %>% 
  group_by(Country) %>%
  summarize(total_deaths = sum(!!sym(last_date))) %>%
  arrange(desc(total_deaths)) %>%
  data.frame()

dat_deaths_long = dat_deaths %>%
  gather(Date, Value, !!sym(first_date):!!sym(last_date)) %>%
  data.frame()

dat_deaths_long$State = NULL
dat_deaths_long$Lat = NULL
dat_deaths_long$Long = NULL

dat_deaths_long = dat_deaths_long %>%
  group_by(Country, Date) %>%
  summarise(total = sum(Value)) %>%
  data.frame()

dat_deaths_long$Date  = as.Date(gsub("_", "-", gsub("date", "", dat_deaths_long$Date)), "%m-%d-%y")

dat_deaths_long = dat_deaths_long %>%
  arrange(Country, Date)

top = 12

```

## Confirmed Cases

Let's look at the daily counts by country. Below are the counts in their natural scale. 

```{r echo=FALSE}
ggplot(subset(dat_cases_long, (Country %in% top_countries$Country[1:top] & total > 100)), 
       aes(x = Date, y = total, color = Country)) + 
  geom_line() + geom_point() + theme_bw(14) + ylab("Number of Confirmed Cases")

```

Below are the top countries by cases as off today, `r Sys.Date()`. 
```{r echo=FALSE}
kable(top_countries[1:15,], caption = "The top countries by cases")
```


Next, let's look at the data on the logarithmic scale. Apart from a few countries, like China, South Korea, Iran, and even Italy, all other countries seem to be very close to a straight line. Suggesting an almost perfect exponential growth in cases.  

```{r echo=FALSE}
ggplot(subset(dat_cases_long, (Country %in% top_countries$Country[1:top] & total > 100)), 
       aes(x = Date, y = total, color = Country)) + 
  geom_line() + geom_point() + theme_bw(14) + ylab("Number of Confirmed Cases") + scale_y_log10()

```

Given how good the exponential fit seems to be, I next perform a regression to predict the future number of cases by country. This is a very simple model, a country specific regression, where we fit a linear regression to the logarithm of the number of cases. In the absence of significant efforts, like the lockdowns that are currently in place, each of the countries below are likely to see million plus cases, in just 10 days. 

```{r echo=FALSE, warning = FALSE}
dat_cases_long_selected = subset(dat_cases_long, (Country %in% c(top_countries$Country[1:top], "India") & total > 100))

fit = lm(log10(total) ~ -1 + Country + Country:Date, data = dat_cases_long_selected,
         weights = )

predict_df = data.frame(expand.grid(Country = unique(dat_cases_long_selected$Country), 
                                    Date = max(dat_cases_long_selected$Date)+1:10))

predict_df$total = 10^predict.lm(fit, newdata = predict_df)
predict_df$type = "Predicted"

dat_cases_long_selected$type = "Observed"

dat_cases_long_selected_predicted = merge(dat_cases_long_selected, predict_df, all = TRUE)

ggplot(subset(dat_cases_long_selected_predicted, Country %in% c("US", "Italy", "Spain","Germany", "India")), 
       aes(x = Date, y = total, color = Country)) + 
  geom_line(aes(linetype= type)) + geom_point(aes(shape = type)) + theme_bw(14) + 
  ylab("Number of Confirmed Cases") + scale_y_log10()

```

### Cases across all countries
Next, let's analyze the number of confirmed cases for all coutries. Again, I fit a simple model to capture the pattern. I fit a degree 3 polynomial to capture the pattern thus far, and extrapolate it for the next 15 days. Additionally, I add a weight that gives greater weights to later time points, the goal of this is to capture the possible effect of the significant worldwide social distancing efforts. Of course, this does not entirely capture the effect of the social distancing, especially what we will see in the coming days. But if nothing were done, this could be a very likely scenario. The scary thing with this fit is that the regression has an Adjusted R-squared of 0.9939, showing how good a fit this simple regression model is to the data.  

```{r echo=FALSE, warning = FALSE}
dat_cases_daily = dat_cases_long %>%
  group_by(Date) %>%
  summarise(daily_total = sum(total)) %>%
  data.frame()

#ggplot(dat_deaths_daily, 
#       aes(x = Date, y = daily_total)) + 
#  geom_line() + geom_point() + theme_bw(14) + ylab("Number of Deaths") + scale_y_log10()

dat_cases_daily$days_from_1stjan = as.numeric(dat_cases_daily$Date - as.Date("2020-01-01"))

# Define a weight function which is quadratic in the number of days. This will be used in the regression to give higher wweights to the later points. 
dat_cases_daily$weights = (dat_cases_daily$days_from_1stjan - min(dat_cases_daily$days_from_1stjan) + 1)^3

fit = lm(log10(daily_total) ~ poly(days_from_1stjan, 3), data = dat_cases_daily, weights = dat_cases_daily$weights)

predict_df = data.frame(days_from_1stjan = max(dat_cases_daily$days_from_1stjan)+1:15,
                        Date = max(dat_cases_daily$Date)+1:15)

predict_df$daily_total = 10^predict.lm(fit, newdata = predict_df)
predict_df$type = "Predicted"

dat_cases_daily$type = "Observed"

dat_cases_daily_predicted = merge(dat_cases_daily, predict_df, all = TRUE)

ggplot(dat_cases_daily_predicted, 
       aes(x = Date, y = daily_total)) + 
  geom_line(aes(linetype= type)) + geom_point(aes(shape = type)) + theme_bw(14) + 
  ylab("Number of Confirmed Cases") + scale_y_log10()


```

## Deaths
Let’s look at the daily counts deaths by country. Below are the counts in their natural scale.
```{r echo = FALSE}
ggplot(subset(dat_deaths_long, (Country %in% top_countries$Country[1:top] & total > 10)), 
       aes(x = Date, y = total, color = Country)) + 
  geom_line() + geom_point() + theme_bw(14) + ylab("Number of Deaths")

```


Below are the top countries by deaths as off today, `r Sys.Date()`. 
```{r echo=FALSE}
kable(top_countries_deaths[1:15,], caption = "The top countries by cases")
```

Let's look at the same plot on the logarithmic scale. 
```{r echo = FALSE}
ggplot(subset(dat_deaths_long, (Country %in% top_countries$Country[1:top] & total > 10)), 
       aes(x = Date, y = total, color = Country)) + 
  geom_line() + geom_point() + theme_bw(14) + ylab("Number of Deaths") + scale_y_log10()

```

We next perform the same excercise as before, fitting a simple linear regression and extrapolating it. These fits are not good for China, and Iran, because these countries have flattened the curve. Even Italy seems to be doing better. But at this rate, the US will have **10,000** deaths within 10 days.  

```{r echo=FALSE, warning = FALSE}
dat_deaths_long_selected = subset(dat_deaths_long, (Country %in% top_countries$Country[1:top] & total > 50))

fit = lm(log10(total) ~ -1 + Country + Country:Date, data = dat_deaths_long_selected)

predict_df = data.frame(expand.grid(Country = unique(dat_deaths_long_selected$Country), 
                                    Date = max(dat_deaths_long_selected$Date)+1:10))

predict_df$total = 10^predict.lm(fit, newdata = predict_df)
predict_df$type = "Predicted"

dat_deaths_long_selected$type = "Observed"

dat_deaths_long_selected_predicted = merge(dat_deaths_long_selected, predict_df, all = TRUE)

ggplot(subset(dat_deaths_long_selected_predicted, Country %in% c("US", "Italy", "Spain","Germany", 
                                                          "Iran", "China", "United Kingdom")), 
       aes(x = Date, y = total, color = Country)) + 
  geom_line(aes(linetype= type)) + geom_point(aes(shape = type)) + theme_bw(14) + 
  ylab("Number of Confirmed Cases") + scale_y_log10()

```

### Deaths across all countries
Next, let's analyze the deaths for all coutries. Again, I fit a simple model to capture the pattern. I fit a degree 3 polynomial to capture the pattern thus far, and extrapolate it for the next 15 days. Of course, this does not capture the effect of the social distancing efforts that are currently on-going all over the world. But if nothing were done, this could be a very likely scenario. The scary thing with this fit is that the regression has an Adjusted R-squared of 0.9986, showing how good a fit this simple regression model is to the data.  

```{r echo=FALSE, warning = FALSE}
dat_deaths_daily = dat_deaths_long %>%
  group_by(Date) %>%
  summarise(daily_total = sum(total)) %>%
  data.frame()

#ggplot(dat_deaths_daily, 
#       aes(x = Date, y = daily_total)) + 
#  geom_line() + geom_point() + theme_bw(14) + ylab("Number of Deaths") + scale_y_log10()

dat_deaths_daily$days_from_1stjan = as.numeric(dat_deaths_daily$Date - as.Date("2020-01-01"))

# Define a weight function which is quadratic in the number of days. This will be used in the regression to give higher wweights to the later points. 
dat_deaths_daily$weights = (dat_deaths_daily$days_from_1stjan - min(dat_deaths_daily$days_from_1stjan) + 1)^3

fit = lm(log10(daily_total) ~ poly(days_from_1stjan, 3), data = dat_deaths_daily, weights = dat_deaths_daily$weights)

predict_df = data.frame(days_from_1stjan = max(dat_deaths_daily$days_from_1stjan)+1:15,
                        Date = max(dat_deaths_long_selected$Date)+1:15)

predict_df$daily_total = 10^predict.lm(fit, newdata = predict_df)
predict_df$type = "Predicted"

dat_deaths_daily$type = "Observed"

dat_deaths_daily_predicted = merge(dat_deaths_daily, predict_df, all = TRUE)

ggplot(dat_deaths_daily_predicted, 
       aes(x = Date, y = daily_total)) + 
  geom_line(aes(linetype= type)) + geom_point(aes(shape = type)) + theme_bw(14) + 
  ylab("Number of Deaths") + scale_y_log10()


```

## Mortality Rate

Next, let's start looking at the mortality rates across countries. We see a big spread, going from 0.5% in Germany to 10% in Italy. 

```{r echo=FALSE, warning = FALSE}
names(top_countries)[2] = "total_cases"
top_countries_both = merge(top_countries, top_countries_deaths, all = TRUE)
#head(top_countries)
top_countries_both$mortality_rate = 100*top_countries_both$total_deaths/top_countries_both$total_cases
top_countries_both$mortality_rate_sd = 100*sqrt(top_countries_both$mortality_rate/100*(1-top_countries_both$mortality_rate/100)/top_countries_both$total_cases)

top_countries_both = top_countries_both %>% arrange(desc(total_cases)) %>% data.frame()

ggplot(top_countries_both[1:top,], aes(x = reorder(Country, mortality_rate), y = mortality_rate, fill = total_cases)) + 
  geom_bar(stat="identity") + geom_errorbar(aes(ymin = (mortality_rate - 2*mortality_rate_sd),
                                                ymax = (mortality_rate + 2*mortality_rate_sd)), width=0.2) + 
  coord_flip() + theme_bw(14) + ylab("Mortality Rate (%)") + xlab("Country")
```

Let's also look at the mortality rate over time (the curves below are smoothed). We see that the mortality rate for most countries seem to be rising. There are two opposing forces:

1. Force 1: The **denominator is low**, not all cases (especially the mild cases are not tested). If this is true, as testing rolls out extensively, we expect to see the mortality rate to fall. 
1. Force 2: The **numerator is low**, many people who die will likely pass away weeks or months from case confirmation. This means that the mortality rate will grow as time passes. 

Let's look at a few cases below. 

* For Italy and Spain, the mortality rate seems to be rising steadily. This is discencerting because the rate is already very high for these counrties. The UK, France and The Netherlands seems to be displaying a very similar pattern. These seem to suggest that we are seeing Force 2 dominate Force 1. The other possibility is that the disease spread is such that the increased testing is not testing enough mild cases. 
* For the US, we see two distinct patterns, first the rate was going down, and then it is going up. Given the lack of testing in the US in the early days, it is likely that the it first went down as many mild cases were detected. 

```{r echo=FALSE, warning = FALSE}
names(dat_cases_long)[3] = "total_cases"
names(dat_deaths_long)[3] = "total_deaths"

dat_both_long = merge(dat_cases_long, dat_deaths_long)
dat_both_long$mortality_rate = dat_both_long$total_deaths/(dat_both_long$total_cases+1) * 100

ggplot(subset(dat_both_long, (Country %in% top_countries_both$Country[1:12] & total_cases > 1000)), 
       aes(x=Date, y = mortality_rate, color=total_cases)) +
  geom_smooth(method = 'loess', span=0.6) + theme_bw() + ylab("Mortality Rate (%)") + 
  facet_wrap(~Country, scales = 'fixed') + theme(axis.text.x=element_text(angle = -90, vjust = 0.5, hjust=1))

```

